Portable Multi- and Many-Core Performance for Finite Difference Codes; Application to the Free-Surface Component of NEMO.

### 2017-06-20 nemolite2d_perf.tex
nemolite-paper: correct collapse and short-loop descriptions. Update gpu perf data

When executing an OpenMP-parallel program on a Non-Uniform Memory
Access (NUMA) compute node it becomes important to ensure that the
memory locations accessed by each thread are local to the hardware
core upon which it is executing. One way of doing this is to implement
a so-called 'first-touch policy' whereby memory addresses that will
generally be accessed by a given thread during program execution are
first initialised by that thread. This is simply achieved by using an
OpenMP-parallel loop to initialise newly-allocated arrays to some
value, \textit{e.g.} zero.

Since data arrays are managed within the GOcean Library this
optimisation can again be implemented without changing the
natural-science code (\textit{i.e.} the Application and Kernel layers).

@@

@@

The Cray compiler is distinct from the other two in that kernel
in-lining does not give any performance benefit and in fact, for the
smaller kernels, it can actually hurt performance. Thus the key
transformation is to encourage the compiler to SIMD vectorise the
Momentum section via a compiler-specific directive (without this it
concludes that such vectorisation would be inefficient). Of the other
transformations, only the addition of the compiler-specific
safe\_address directive (see Section~\ref{sec_safe_address}) was found
to give any benefit.

\begin{table}[!t]
  \caption{Performance (millions of points updated per second) on an Intel
    Ivy Bridge CPU for the
  $256^2$ case after each code transformation. Where an optimisation
  utilises compiler-specific directives then performance figures for
  the other compilers are omitted.}
\label{TABLE_opt_breakdown}
\centering
\begin{tabular}{l|c|c|c}
\hline
Compiler:           & Gnu & Intel & Cray  \\
\hline
Original         & 3.87 & 12.2 & 6.83 \\
Vanilla {PS}y{KA}l  & 3.07 & 6.55 & 6.73 \\
Safe-address     & --   & --   & 6.95 \\
In-line Momentum & 4.31 & 10.7 & 6.93 \\
SIMD Momentum    & --   & --   & 11.8 \\
\parbox{2.5cm}{\raggedright Grid data from read-only objects} & 4.31 & 11.3 & 11.8 \\
In-line Continuity         & 4.83 & 11.8 & 11.6 \\
In-line remaining kernels  & 5.89 & 12.0 & 11.5 \\
In-line field copies       & 5.92 & 12.5 & 11.4 \\
\hline
\parbox{2.5cm}{\raggedright \%-speed-up of best \textit{c.f.} original} & 34.6 & 2.39 & 42.2 \\
\hline
\end{tabular}
\end{table}

@@

\subsubsection{OpenACC Version}

The advantage of the {PS}y{KA}l re-structuring becomes apparent if we
wish to run NEMOLite2D on different hardware, \textit{e.g.} a
GPU. This is because the necessary code modifications are, by design,
limited to the middle PSy layer. In order to demonstrate this and to
check for any limitations imposed by the {PS}y{KA}l re-structuring, we
had a GPU expert from NVIDIA port the Fortran NEMOLite2D to GPU. He
used OpenACC directives for this porting as that approach is similar
to the use of OpenMP directives and works well within the {PS}y{KA}l
apprach. He also experimented with using CUDA directly within the
original form of NEMOLite2D in order to quantify any performance
penalty incurred by taking the {PS}y{KA}l/OpenACC approach.

>>>>

When executing an OpenMP-parallel program on a Non-Uniform Memory
Access (NUMA) compute node it becomes important to ensure that the
memory locations accessed by each thread are local to the hardware
core upon which it is executing. One way of doing this is to implement
a so-called `first-touch policy' whereby memory addresses that will
generally be accessed by a given thread during program execution are
first initialised by that thread. This is simply achieved by using an
OpenMP-parallel loop to initialise newly-allocated arrays to some
value, \textit{e.g.} zero.

Since data arrays are managed within the GOcean infrastructure this
optimisation can again be implemented without changing the
natural-science code (\textit{i.e.} the Application and Kernel layers).

@@

\subsection{Construction of OpenACC-Parallel NEMOLite2D}
\label{sec_oacc_steps}

The advantage of the {PS}y{KA}l re-structuring becomes apparent if we
wish to run NEMOLite2D on different hardware, \textit{e.g.} a
GPU. This is because the necessary code modifications are, by design,
limited to the middle PSy layer. In order to demonstrate this and to
check for any limitations imposed by the {PS}y{KA}l re-structuring, we
had an expert from NVIDIA port the Fortran NEMOLite2D to GPU. He
used OpenACC directives for this porting as that approach is similar
to the use of OpenMP directives and works well within the {PS}y{KA}l
approach.

\subsubsection{Data Movement}

Although the advent of technologies such as NVLink are alleviating the
bottleneck presented by the connection of the GPU to the CPU, it
remains critical to minimise data movement between the memory spaces
of the two processing units. In NEMOLite2D this is achieved by
performing all computation on the GPU. The whole time-stepping loop
can then be enclosed inside a single OpenACC \texttt{data} region and it
is only necessary to bring data back to the CPU for the purposes
of I/O.

\subsubsection{Kernel Acceleration}

Moving the kernels to execute on the GPU was achieved by using the
OpenACC \texttt{kernels} directive in each Fortran routine containing
loops over grid points. (In the {PS}y{KA}l version this is just the
single {PS}y layer subroutine). This directive instructs the OpenACC
compiler to automatically create a GPU kernel for each loop nest it
encounters.

\subsubsection{Force parallelisation of Flather kernels}

The PGI compiler was unable to determine whether the loops applying
the Flather boundary condition in the $x$ and $y$ directions were safe
to parallelise. This information therefore had to be supplied by
inserting a \texttt{loop independent} OpenACC directive before each of
the two loops.

\subsubsection{Loop collapse}

When using the OpenACC \texttt{kernels} directive the compiler creates
GPU kernels by parallelising only the outer loop of each loop nest.
For the majority of loop nests in NEMOLite2D all of the iterations are
independent and the loop bodies consist of just a handful of
executable statements. For small kernels the loop start-up cost can be
significant and therefore it is beneficial to generate as many threads
as will fit on the GPU and then re-use them as required (\textit{i.e.} if the
data size is greatre than the number of threads). For this approach to
be efficient we must expose the maximum amount of parallelism to the
GPU and we do this by instructing the compiler to parallelise both
levels (i.e. \texttt{ji} and \texttt{jj}) of a loop nest. This is
achieved using the OpenACC \texttt{loop collapse(2)} directive for all
of the smaller kernels in NEMOLite2D.

\subsubsection{Shortloop}

In contrast to all of the other NEMOLite2D kernels, the Momentum
kernels ($u$ and $v$) are relatively large in terms of the number of
executable statements they contain. In turn, this means that the
corresponding kernels will require more state and thus more registers
on the GPU. Therefore, rather than generating as many threads as will
fit on the GPU, it is more efficient only to generate as many as
required by the problem so as to minimise reigster pressure. Once
slots become free on the GPU, new threads will launch with the
associated cost amortized by the longer execution time of the larger
kernel. The compiler can be persuaded
to adopt the above strategy through the use of the (PGI-specific)
\texttt{loop shortloop} directive which tells it that there is not
likely to be much thread re-use in the following loop. This directive
was applied to both the \texttt{ji} and \texttt{jj} loops in both of
the Momentum kernels.

\subsubsection{Kernel fusion}

Both of the Momentum kernels read from 20 double-precision arrays and
thus require considerable memory bandwidth.  However, the majority of
these arrays are used by both the $u$ and $v$ kernels. It is therefore
possible to reduce the memory-bandwidth requirements by fusing the two
kernels together. In practice, this means fusing the two,
doubly-nested loops so that we have a single loop nest containing both
the $u$ and $v$ kernel bodies/calls.

\subsubsection{CUDA}

We also experimented with using CUDA directly within the original form
of NEMOLite2D in order to quantify any performance penalty incurred by
taking the {PS}y{KA}l/OpenACC approach.

@@

The Cray compiler is distinct from the other two in that kernel
in-lining does not give any performance benefit and in fact, for the
smaller kernels, it can actually hurt performance. Thus the key
transformation is to encourage the compiler to SIMD vectorise the
Momentum section via a compiler-specific directive (without this it
concludes that such vectorisation would be inefficient). Of the other
transformations, only the change to constant loop bounds and the
addition of the compiler-specific safe\_address directive (see
Section~\ref{sec_safe_address}) were found to (slightly) improve
performance.

\begin{table}[!t]
  \caption{Performance (millions of points updated per second) on an Intel
    Ivy Bridge CPU for the
  $256^2$ case after each code transformation. Where an optimisation
  utilises compiler-specific directives then performance figures for
  the other compilers are omitted.}
\label{TABLE_opt_breakdown}
\centering
\begin{tabular}{l|c|c|c}
\hline
Compiler:            & Gnu & Intel & Cray  \\
\hline
Original             & 3.87 & 12.2 & 6.83 \\
Vanilla {PS}y{KA}l   & 2.59 & 6.27 & 6.35 \\
Constant loop bounds & 3.07 & 6.55 & 6.73 \\
Safe-address         & --   & --   & 6.95 \\
In-line Momentum     & 4.31 & 10.7 & 6.93 \\
SIMD Momentum        & --   & --   & 11.8 \\
\parbox{2.5cm}{\raggedright Grid data from read-only objects} & 4.31 & 11.3 & 11.8 \\
In-line Continuity         & 4.83 & 11.8 & 11.6 \\
In-line remaining kernels  & 5.89 & 12.0 & 11.5 \\
In-line field copies       & 5.92 & 12.5 & 11.4 \\
\hline
\parbox{2.5cm}{\raggedright \%-speed-up of best \textit{c.f.} original} & 34.6 & 2.39 & 42.2 \\
\hline
\end{tabular}
\end{table}

@@

\subsubsection{OpenACC}

The advantage of the {PS}y{KA}l re-structuring becomes apparent if we
wish to run NEMOLite2D on different hardware, \textit{e.g.} a
GPU. This is because the necessary code modifications are, by design,
limited to the middle PSy layer. In order to demonstrate this and to
check for any limitations imposed by the {PS}y{KA}l re-structuring, we
had an expert from NVIDIA port the Fortran NEMOLite2D to GPU. He
used OpenACC directives for this porting as that approach is similar
to the use of OpenMP directives and works well within the {PS}y{KA}l
apprach. He also experimented with using CUDA directly within the
original form of NEMOLite2D in order to quantify any performance
penalty incurred by taking the {PS}y{KA}l/OpenACC approach.
